---
title: "Tutorial2_NHANES_data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial2_NHANES_data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  fig.align = "center",
  fig.height = 6,
  fig.width = 6,
  comment = "#>"
)
```

```{r setup, message = FALSE}
# loading packages
library(mvnimpute)
library(dplyr)
```

In this tutorial, we will use a small subset of PCB variables from the NHANES 2011 - 2012 data set in this package. We will start from the data preparation step to the multiple imputation step to show how can we prepare any arbitrary data to the required structure that can be used for multiple imputation using this packages. First of all, let us see the structure of the data

```{r, echo = FALSE}
# 2011-2016 NHANES data
# measurement data
dat <- nhanes.dat$data %>%
  select("SDDSRVYR", starts_with("LBC"), -ends_with("LA"))
# censoring indicator
dat.ind <- nhanes.dat$data %>%
  select("SDDSRVYR", starts_with("LBD"))
# censoring values
dat.censor <- as.matrix(nhanes.dat$PCB.LOD[1, ][1:10])

# transform data frames to matrices
dat <- as.matrix(subset(dat, SDDSRVYR == 7))
dat.ind <- as.matrix(subset(dat.ind, SDDSRVYR == 7))
# recode missing values to 0
dat.ind <- ifelse(is.na(dat.ind), 0, dat.ind)

# select first 10 variables
dat <- dat[, 2:11]
dat.ind <- dat.ind[, 2:11]

# remove unit nonresponse
unit.non <- which(apply(is.na(dat), 1, sum) == 10)
dat <- dat[-unit.non, ]


# rename 
check.names <- sum(substr(colnames(dat), 4, 6) == substr(colnames(dat.ind), 4, 6) & substr(colnames(dat), 4, 6) == substr(colnames(dat.censor), 4, 6))
if (check.names == 10) colnames(dat) <- colnames(dat.ind) <- colnames(dat.censor) <- paste0("PCB", substr(colnames(dat), 4, 6))
```

Since the measurements data can not be negative, we take log transformation to cirvumvent negative values in imputation.

```{r}
# log transformation
dat <- log(dat)
dat.censor <- log(dat.censor)
```

## 1. Summary of the data

We see that there are 10 variables, and the sample size is 3168

```{r, echo = FALSE}
## data structures
dims <- dim(dat)
n <- dims[1]
p <- dims[2]
cat("Number of observations: ", n, "\n")
cat("Number of variables: ", p)
```
```{r, echo = FALSE}
## summary of each variable
summary(dat)
```

The table below shows the first few lines of the log-transformed data, we observe that the missing values are all from unit-nonrepsonse. After removing those unit-nonresponse that may create problems in the convergence of the multiple imputation algorithm, there are only cneosred values left in the resulting data.

```{r, echo = FALSE}
head(dat)
```

The vector below shows the log-transformed limits of detection (LOD) for each variable. In this case, the censoring is **left censoring** since the values fall below the respective limits of detection are censored.

```{r, echo = FALSE}
head(dat.censor)
```

## 2. Data preparation

Next, we will perform some preliminary work so that the data can be transformed in the required format that can be used for the multiple imputation algorithm in this package.

In order to visualize the percentages of the missing, observed data censored values in the dataset, we need to create a data type indicator matrix, in which 0, 1, and 2 are for missing, observed and censored values, respectively.

```{r}
## creat a matrix for including the new data type indicators
new.ind <- dat.ind[-unit.non, ]
## recode censoring index to 2
new.ind <- ifelse(new.ind == 1, 2, new.ind)
## recode observed index to 1
new.ind <- ifelse(!is.na(dat) & new.ind != 2, 1, new.ind)
```

The table below shows the first few lines of the newly created data type indicator matrix.

```{r}
head(new.ind)
```

As explained in the other vignette, we assign a pair of values to each of the data point which we will use to check the type of data in the multiple imputation algorithm. Thus, we will need to create a list of two that includes those pairs. We name the list to be created as `incomplete.data`, as it contains the information of unobserved values.

```{r}
## create the list of two 
incomplete.data <- list()
## assign the initial matrices to contain NA values
incomplete.data[[1]] <- incomplete.data[[2]] <- matrix(NA, nrow = nrow(new.ind), ncol = ncol(new.ind))

for (i in 1:n) {
  for (j in 1:p) {
    # observed values
    if (new.ind[i, j] == 1) {
      incomplete.data[[1]][i, j] <- incomplete.data[[2]][i, j] <- dat[i, j]
    }
    # missing values
    else if (new.ind[i ,j] == 0) {
      incomplete.data[[1]][i, j] <- -10e10
      incomplete.data[[2]][i, j] <- 10e10
    }
    # censored values: left censoring
    else if (new.ind[i, j] == 2) {
      incomplete.data[[1]][i, j] <- -10e10
      incomplete.data[[2]][i ,j] <- dat[i, j]
    }
  }
}

## rename variable name in each of the matrix in the list
colnames(incomplete.data[[1]]) <- colnames(incomplete.data[[2]]) <- colnames(dat)
```

The table below show the first few lines the two newly created matrices, repectively.

```{r}
## lower limits
head(incomplete.data[[1]])
## upper limits
head(incomplete.data[[2]])
```

## 3. Data visualization

`visual.plot` only required the data type indicator matrix as the input.

```{r}
visual.plot(new.ind)
```

## 4. Multiple imputation

We will be ready to run multiple imputation for the incomplete data after specifying the prior parameters for the Normal-Inverse-Wishart distribution.

### 4.1 Prior specifications and starting values

We have to exactly follow the format of the list of prior specifications as described in the other vignette.

```{r}
## calculate the CC and AC parameters
params <- param.calc(incomplete.data)
## prior specifications
prior <- list(
  mu.0 = params$AC.mean,
  Lambda.0 = params$AC.var,
  kappa.0 = 100,
  nu.0 = p * (p + 1) / 2
)

## starting values
starting <- list(
  mu = params$CC.mean,
  sigma = params$CC.cov
)
```

### 4.2 Multiple imputation

The *multiple.imputation* function requires five arguments:

+ The list of two generated in the data generation step that include the missing and censored information
+ The prior specifications
+ The starting values
+ The rounds of running multiple imputation
+ The logical indicator whether to print out the running status

```{r, message = FALSE}
iter <- 500
sim.dat <- multiple.imputation(incomplete.data,
                               prior.params = prior, 
                               starting.values = starting, 
                               iter = iter, 
                               details = TRUE)
```

## 5. Diagnostic graphs

### 5.1 Simulated mean values

```{r}
conv.plot(sim.dat$simulated.mu, iter)
```

### 5.2 Simulated variance values

```{r}
conv.plot(sim.dat$simulated.sig, iter)
```

### 5.3 Autocorrelation functions for the mean

```{r}
acf.calc(sim.dat$simulated.mu, title = colnames(sim.dat$simulated.mu))

```

### 5.4 Autocorrelation functions for the variance

```{r}
acf.calc(sim.dat$simulated.sig, title = colnames(sim.dat$simulated.sig))
```
